An Empirical Study of Invasion and Resistance for Iterated Prisoner's Dilemma Strategies

### 2017-05-15 main.tex
A variety of edits for the paper

% Title
\title{An Empirical Study of Invasion and Resistance
       for Iterated Prisoner's Dilemma Strategies}
\author{Vincent Knight \and
        Marc Harper \and
        Nikoleta E. Glynatsi \and
        Owen Campbell} % TODO Authors?
\date{}

@@

\begin{abstract}
    The Iterated Prisoner's Dilemma is a well established framework for
    the study of emergent behaviour. In this paper an extensive numerical
    study of the evolutionary dynamics of this framework are presented. Fixation
    probabilities for Moran processes are obtained for 164
    different strategies. It is found that players with long memories and
    sophisticated behaviours outperform many strategies that perform well
    in a two player setting.
\end{abstract}

@@

The Prisoner's Dilemma (PD) \cite{Flood1958} is a fundamental two player game
used to model a large variety of strategic interactions. Each player can choose
between cooperation (C) or defection (D). The decisions are made simultaneously
and independently. The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This extended form is called the Iterated Prisoner's
Dilemma (IPD). As described in \cite{Axelrod1980a, Knight2016, Press2012} a
number of strategies have been developed to take advantage of the history of
play. Recently, some strategies referred to as Zero Determinant strategies
\cite{Press2012} even manipulate players through extortion.

The Moran Process \cite{Moran1957} is a model of
evolutionary population dynamics that has been used to gain insights about the
evolutionary stability in a number of settings (more details given in
Section~\ref{sec:the_moran_process}).
Several earlier
works have studied iterated games in the context of the prisoner's dilemma
\cite{Nowak, Stewart2012}, however these often make simplifying assumptions
and/or do not consider sophisticated behaviour: only considering strategies that
either cooperate or defect.

This manuscript provides a detailed numerical analysis of
\textbf{\input{./tex/num_strategies.tex}}complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over \input{./tex/num_strategies_axelrod.tex}parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, Zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: same use
knowledge of the game such as the number of turns and others have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 40 contributions made by
different programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in Section~\ref{sec:strategies}.

The library can conduct matches, tournaments and population
dynamics with variations including noise and spatial structure.
The strategies and simulation frameworks are
automatically tested to an extraordinarily high degree of coverage in accordance
with best research software practices.

@@

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform intelligently designed
     strategies.
 \item Strong resistors specifically evolve or have a handshake mechanism. This
     has the potential to offer insight in to organisms such as antibacterial
     resistant bacteria \cite{Davies2010}.
\end{enumerate}

@@

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process, a stochastic birth death process on a finite population in which the
population size stays constant over time. Individuals are \textbf{selected}
according to a given fitness landscape. Once selected, the individual is
reproduced and similarly another individual is chosen to be removed from the
population. In some settings mutation is also considered but without mutation
(the case considered in this work) this process will arrive at an absorbing
state where the population is entirely made up of players of one strategy. The
probability with which a given strategy is the survivor is called the
\textit{fixation probability}. A more detailed analytic description of this is
given in Section~\ref{sec:validation}.

@@

A part from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, all these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.

More
information about each player can be obtained in the documentation for
\cite{axelrodproject} and a detailed description of the performance
 of these strategies in IPD
tournaments will be described in upcoming manuscript.
% TODO Think of how to reference this upcoming manuscript

All of the training code is available in the
Axelrod repository with documentation to train new strategies easily. Training
% TODO Include link etc
typically takes less than 200 iterations and can be completed within several
hours on commodity hardware.

@@

\begin{itemize}
	\item TF1 \(N=12\), 0\% noise, 10000 repetitions per match
	\item TF2 \(N=10\), 0\% noise, 10000 repetitions per match
	\item TF3 \(N=8\), 1\% noise, 100 repetitions per match
\end{itemize}

@@

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective. Interestingly it has no handshake
mechanism, this could be due to the fact that it is trained in a noisy
environment.

@@

TF1 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF1 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{prison}) manages to
achieve cooperation with TF1 after about 20 rounds of play.

@@

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken
from a large number of match outcomes for the pairs of players for use in
computing fitnesses in the Moran process. This approach was verified to agree
with unsampled calculations to a high degree of accuracy in specific cases.
This is described in Algorithms~\ref{alg:data_collection}
and~\ref{alg:moran_process}..

@@

\begin{equation}\label{equ:expected_payoff_one}
    F_i = \frac{a(i - 1) + b(N - i)}{N - 1}
\end{equation}

\begin{equation}\label{equ:expected_payoff_two}
    G_i = \frac{ci + d(N - i - 1)}{N - 1}
\end{equation}

With an intensity of selection \(\omega\) the fitness of both strategies is
given by:

\begin{equation}\label{equ:expected_payoff_one}
    f_i = 1 - \omega + \omega F_i
\end{equation}

\begin{equation}\label{equ:expected_payoff_two}
    g_i = 1 - \omega + \omega G_i
\end{equation}

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match up between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategies pairs.

@@

Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of the analytical formulae that relies on the average payoffs
(\ref{equ:payoff_matrix}).

All data generated for this validation exercise can be found
at \cite{data}.

A detailed analysis of the 164 strategies considered,
using direct Moran processes will be shown in the next Section.

@@

When $N=2$ the Moran process is effectively a measure of the relative
mean payoffs over all possible matches between two players. The strategy
that scores higher than the other more often will fixate more often.

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism (a cooperation followed by a defection on the first
        move). As long as the opponent plays the same handshake and does not
        defect in the future it cooperates. Otherwise it defects for all
        rounds. This strategy was specifically designed for
        evolutionary processes \cite{Li2009}.
    \item The Defector: it always defects. As it has little potential
        interaction with itself (recall that \(N=2\)), its
        aggressiveness is rewarded.
    \item The Aggravater strategy which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake: a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself: just as the Collective strategy does. However it
        is not aggressive enough to invade other populations (as can be seen in
        Section~\ref{sec:strong_invaders}.
\end{enumerate}

@@

\begin{table}[!hbtp]
    \centering
    \input{./tex/summary_top_2_invade.tex}
    \caption{Summary of top five strategies for \(N=2\)}
    \label{tbl:summary_top_2}
\end{table}

As will be demonstrated in Section~\ref{sec:population_size} the results for
\(N=2\) differ from those of larger $N$. Hence these results do not concur with
the literature which suggests that Zero Determinant strategies should be
effective for larger population sizes, but these analysis consider stationary
behaviour, while this work runs for a fixed number of rounds. Note that the
stationarity assumptions allows for the analysis to take place leading to the
conclusions about zero determinant strategies however it is dependent on a
specific structure of strategies. The analysis carried out here makes no
assumptions about the structure of the strategies by using actual simulation
interactions.

In the next sections close attention to
strategies who are strong invaders/resistors is given.

@@

In this section the focus is on the ability of a mutant strategy to invade: the
probability of 1 individual of a given type successfully becoming fixated in a
population of \(N - 1\) other individuals, denoted by \(x_1\). The fixation
probabilities are shown in
Figures~\ref{fig:boxplot_3_invade},~\ref{fig:boxplot_7_invade}
and~\ref{fig:boxplot_14_invade} for \(N\in\{3, 7, 14\}\) showing the mean
fixation as well as the neutral fixation for each given scenario.

@@

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic although close inspection of the
source code of PSO Gambler shows that it makes stochastic decisions with extreme
probabilities: almost always acting as expected. Apart from TF3 in \(N=3\), the
finite state machines trained specifically for Moran processes do not appear in
the top 5: whereas strategies trained for tournament do. This is due to the
nature of invasion: most of the opponents will initially be different
strategies. The next section will consider the converse situation.

@@

Interestingly none of these strategies are stochastic: this is explained by
the need of strategies to have a steady hand when interacting with their own
kind. In essence: acting stochastically increase the chance of friendly fire.
However it is possible to design a strategy with a ``stochastic
handshake'' \cite{Lee2015}.

There are are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF1 and TF2. These two strategies are with CS the strongest
resistors. They all have handshakes, and whilst the handshake that CS and
Handshake (which ranks highly for the smaller values of \(N\)) have was
programmed, the handshakes of TF1 and TF2 evolved through an evolutionary
process.

@@

The payoff maximizing strategies typically will not defect before the opponent's
first defection, possibly because the training strategy collection contains a
significant portion of strategies such as Grudger and Fool Me Once that
retaliate harshly by defecting for the remainder of the match if the opponent
has more than a small number of cumulative defections. Paradoxically it is
necessary to defect (as a signal) in order to achieve mutual cooperation with
opponents using the same strategy but not with other opponents.

A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate first round defections. So the
handshakes used by TF1 and TF2, and CS, are in some sense optimal. These
discoveries may have significant ramifications regarding the evolution of
cooperation and forgiveness in biological organisms such as antibacterial
resistant bacteria and social interactions
between humans.

@@

This is confirmed by the performance of the Zero determinant strategies, whilst
some do rank relatively highly for \(N=2\) (ZD-extort-4 has tank 16) this rank
does not translate to larger populations.

@@

@@

When studying evolutionary processes it is vital to consider \(N>2\) as the
special case for \(N=2\) cannot be used to extrapolate performance in bigger
populations. This was shown both observationally in
Sections~\ref{sec:strong_invaders} and~\ref{sec:strong_resistors} but also by
considering the correlation of the ranks in different population sizes in
Section~\ref{sec:population_size}.

Memory one strategies do not perform well, as predicted by \cite{Press2012}.
There are no memory one strategies in the top 5 performing strategies
for \(N>3\). This is due to their lack of sophistication which allows them to
recognise and adjust to their opponent. Some very sophisticated strategies
proves to be high performers for invasion: these are infinite memory strategies
which have been trained using a number of reinforcement learning algorithms.
Interestingly they have been trained to perform well in tournaments and not
Moran processes which highlights the potentially for improvement.

One of the major findings discussed in Section~\ref{sec:strong_resistors}, is
the ability of strategies with a handshake mechanism to resist invasion. This
was not only identify for CS (an intelligently designed strategy) but also for
two FSM strategies (TF1 and TF2) specifically trained through an evolutionary
process. In these two cases, the handshake mechanism was a product of the
evolutionary process. This has the potential to help with the understanding of
organisms with a strong resistance to invasion such as anti antibacterial
resistant bacteria \cite{Davies2010}. With the knowledge that a handshake being
likely to exist, perhaps it can be mimicked.

It is felt that these findings are important for the ongoing understanding of
population dynamics and offer evidence for some of the shortcomings of short
memory which has started to be recognised by the community \cite{Hilbe2017}.

@@

There are various areas for further work to build on this. Firstly, an analysis
of the effect of noise would offer insights about the stability of the findings.
It would also be possible to consider three or more types of strategy in the
population and finally mutation would also offer an interesting dimension to
explore.

>>>>

% Title
\title{An Empirical Study of Invasion and Resistance
       for Iterated Prisoner's Dilemma Strategies}
\title{Evolving Invaders and Resistors for the Iterated Prisoner's Dilemma}
\author{Vincent Knight \and
        Marc Harper \and
        Nikoleta E. Glynatsi \and
        Owen Campbell} % TODO Authors?
\date{}

@@

\begin{abstract}
    The Iterated Prisoner's Dilemma is a well established framework for
    the study of emergent behaviour. In this paper an extensive numerical
    study of the evolutionary dynamics of this framework are presented. Fixation
    probabilities for Moran processes are obtained for 164
    different strategies. It is found that players with long memories and
    sophisticated behaviours outperform many strategies that perform well
    in a two player setting. Moreover we introduce several strategies
    trained with evolutionary algorithms to excel at the Moran process.
    These strategies are excellent invaders and resistors of invasion.
\end{abstract}

@@

The Prisoner's Dilemma (PD) \cite{Flood1958} is a fundamental two player game
used to model a large variety of strategic interactions. Each player can choose
between cooperation (C) or defection (D). The decisions are made simultaneously
and independently. The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This extended form is called the Iterated Prisoner's
Dilemma (IPD). As described in \cite{Axelrod1980a, Knight2016, Press2012} a
number of strategies have been developed to take advantage of the history of
play. Recently, some strategies referred to as Zero Determinant strategies
\cite{Press2012} can manipulate some players through extortionate mechanisms.

The Moran Process \cite{Moran1957} is a model of
evolutionary population dynamics that has been used to gain insights about the
evolutionary stability in a number of settings (more details given in
Section~\ref{sec:the_moran_process}).
Several earlier
works have studied iterated games in the context of the prisoner's dilemma
\cite{Nowak, Stewart2012}, however these often make simplifying assumptions
and/or do not consider sophisticated behaviour: only considering strategies that
either cooperate or defect, or are limited to classes of strategies
such as memory-one strategies that only use the previous round of play.

This manuscript provides a detailed numerical analysis of
\textbf{\input{./tex/num_strategies.tex}}complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over \input{./tex/num_strategies_axelrod.tex}parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, Zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: excluded
are those that make use of knowledge of the game such as the number of turns
and others that have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 40 contributions made by
different programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in Section~\ref{sec:strategies}.

In addition to providing a large collection of strategies, the Axelrod library
can conduct matches, tournaments and population
dynamics with variations including noise and spatial structure.
The strategies and simulation frameworks are
automatically tested to an extraordinarily high degree of coverage in accordance
with best research software practices.

@@

\begin{enumerate}
 \item Zero determinant strategies are not particularly effective for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform designed strategies.
 \item Strong resistors specifically evolve or have a handshake mechanism.
\end{enumerate}

These finding have the potential to offer insight in to organisms such as antibacterial
resistant bacteria \cite{Davies2010} and human social behaviors.

@@

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process, a stochastic birth death process on a finite population in which the
population size stays constant over time. Individuals are \textbf{selected}
according to a given fitness landscape. Once selected, the individual is
reproduced and similarly another individual is chosen to be removed from the
population. In some settings mutation is also considered but without mutation
(the case considered in this work) this process will arrive at an absorbing
state where the population is entirely made up of players of one strategy. The
probability with which a given strategy is the survivor is called the
\textit{fixation probability}. A more detailed analytic description of this is
given in Section~\ref{sec:validation}. In our simulations offspring do not
inherit any knowledge or history from parent replicants.

@@

A part from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were not
trained specifically for the Moran process or to cooperate well with themselves
(though they all do). Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.

More information about each player can be obtained in the documentation for
\cite{axelrodproject} and a detailed description of the performance
 of these strategies in IPD tournaments will be described in upcoming manuscript(s).
% TODO Think of how to reference this upcoming manuscript

All of the training code is available in the
Axelrod repository with documentation to train new strategies easily. Training
% TODO Include link etc
typically takes less than 100 generations and can be completed within several
hours on commodity hardware.

@@

\begin{itemize}
	\item TF1 \(N=12\), 0\% noise, 10000 repetitions per matchup
	\item TF2 \(N=10\), 0\% noise, 10000 repetitions per matchup
	\item TF3 \(N=8\), 1\% noise, 100 repetitions per matchup
\end{itemize}

@@

TF3 cooperates and defects with
various cycles depending on the opponent's actions. TF3 will mutually
cooperate with any strategy and only tolerates a few defections before
defecting for the rest of match. It is similar to but not exactly the same as
Fool Me Once, a strategy that cooperates until the opponent has defected twice
(not necessarily consecutively), and defects indefinitely thereafter. Though a
product of training with a Moran objective. It differs from TF1 and TF2 
by lacking a handshake mechanism.

@@

TF1 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF1 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{prison}) manages to
achieve cooperation with TF1 after about 20 rounds of play. TF1 is functionally
very similar to a strategy known as ``Collective Strategy'', which has a
handshake of CD and cooperates with opponents that matched the handshake
until they defect, defecting thereafter if the opponent ever defects \cite{Li2009}.
This strategy was specifically designed for evolutionary processes.

@@

Each strategy pair is run for 1000 repetitions of the Moran process to fixation
with starting population distributions of $(1, N-1)$, $(N/2, N/2)$ and $(N-1 ,
1)$, for \(N\) from 2 through 14. The fixation probability is then empirically
computed for each combination of starting distribution and value of \(N\).  The
axelrod library can carry out exact simulations of the Moran process. Since some
of the strategies have a high computational cost or are stochastic, samples are
taken from a large number of match outcomes for the pairs of players for use in
computing fitnesses in the Moran process. This approach was verified to agree
with unsampled calculations to a high degree of accuracy in specific cases.
This is described in Algorithms~\ref{alg:data_collection}
and~\ref{alg:moran_process}.

@@

\begin{equation}\label{equ:expected_payoff_one}
    f_i = \frac{a(i - 1) + b(N - i)}{N - 1}
\end{equation}

\begin{equation}\label{equ:expected_payoff_two}
    g_i = \frac{ci + d(N - i - 1)}{N - 1}
\end{equation}

% With an intensity of selection \(\omega\) the fitness of both strategies is
% given by:
% 
% \begin{equation}\label{equ:expected_payoff_one}
%     f_i = 1 - \omega + \omega F_i
% \end{equation}
% 
% \begin{equation}\label{equ:expected_payoff_two}
%     g_i = 1 - \omega + \omega G_i
% \end{equation}

@@

Comparisons of \(x_1, x_{N/2}, x_{N-1}\) are shown in
Figure~\ref{fig:comparison_deterministic}. The points represent the simulated
values and the line shows the theoretical value. Note that these are all
deterministic strategies and show a perfect match up between the expected value
of (\ref{equ:fixation_probability}) and the actual Moran process for all
strategies pairs. Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of the analytical formulae that relies on the average payoffs
(\ref{equ:payoff_matrix}). Estimating all the transitions may give better agreement
but is not necessary for our purposes. All data generated for this validation exercise can be found
at \cite{data}.

@@

@@

When $N=2$ the Moran process is effectively a measure of the distribution of relative
mean payoffs over all possible matches between two players. The strategy
that scores higher than the other more often will fixate more often.

@@

\begin{enumerate}
    \item The top strategy is the Collective Strategy (CS) which has a simple
        handshake mechanism described above.
    \item The Defector: it always defects. As it has little potential
        interaction with itself (recall that \(N=2\)), its
        aggressiveness is rewarded.
    \item The Aggravater strategy which plays like Grudger (responding to any
        defections with unconditional defections throughout) however starts by
        playing 3 defections.
    \item Predator, a finite state machine described in \cite{Ashlock2006}.
    \item Handshake: a slightly less aggressive version of the Collective
        strategy \cite{robson1989}. As long as the initial sequence is played
        then it cooperates. Thus it will do well in a population consisting of
        many members of itself: just as the Collective strategy does. However it
        is not aggressive enough to invade other populations (as can be seen in
        Section~\ref{sec:strong_invaders}.
\end{enumerate}

@@

\begin{table}[!hbtp]
    \centering
    \input{./tex/summary_top_2_invade.tex}
    \caption{Summary of top five strategies for \(N=2\) (neutral fixation is $p=0.5$.}
    \label{tbl:summary_top_2}
\end{table}

As will be demonstrated in Section~\ref{sec:population_size} the results for
\(N=2\) differ from those of larger $N$. Hence these results do not concur with
the literature which suggests that Zero Determinant strategies should be
effective for larger population sizes, but these analysis consider stationary
behaviour, while this work runs for a fixed number of rounds. Note that the
stationarity assumptions allows for the analysis to take place leading to the
conclusions about zero determinant strategies however it is dependent on a
specific structure of strategies. The analysis carried out here makes no
assumptions about the structure of the strategies by using actual simulation
interactions. \cite{Stewart...}

% In the next sections close attention to
% strategies who are strong invaders/resistors is given.

@@

In this section the focus is on the ability of a mutant strategy to invade: the
probability of one individual of a given type successfully fixating in a
population of \(N - 1\) other individuals, denoted by \(x_1\). The fixation
probabilities are shown in
Figures~\ref{fig:boxplot_3_invade},~\ref{fig:boxplot_7_invade}
and~\ref{fig:boxplot_14_invade} for \(N\in\{3, 7, 14\}\) showing the mean
fixation as well as the neutral fixation for each given scenario.

@@

As well as noting that the memory length and complexity of these strategies are
much greater than one, it is interesting to note that none of them are akin to
memory one strategies. Only one is stochastic although close inspection of the
source code of PSO Gambler shows that it makes stochastic decisions rarely, and
is functionally very similar to its deterministic cousing EvolvedLookerUp. Apart
from TF3 in \(N=3\), the finite state machines trained specifically for Moran
processes do not appear in the top 5: whereas strategies trained for tournament do.
This is due to the nature of invasion: most of the opponents will initially be different
strategies. The next section will consider the converse situation.

@@

Interestingly none of these strategies are stochastic: this is explained by
the need of strategies to have a steady hand when interacting with their own
kind. In essence: acting stochastically increases the chance of friendly fire.
However it is possible to design a strategy with a stochastic or error-correcting
handshake that is an excellent resistor even in noisy environments \cite{Lee2015}.

There are are only two new strategies that appear in the top ranks for
\(x_{N-1}\): TF1 and TF2. These two strategies are with CS the strongest
resistors. They all have handshakes, and whilst the handshakes of CS and
Handshake (which ranks highly for the smaller values of \(N\)) were
programmed, the handshakes of TF1 and TF2 evolved through an evolutionary
process without any priming.

@@

The payoff maximizing strategies typically will not defect before the opponent's
first defection, possibly because the training strategy collection contains some
strategies such as Grudger and Fool Me Once that retaliate harshly by defecting
for the remainder of the match if the opponent has more than a small number of
cumulative defections. Paradoxically it is advantageous to defect (as a signal)
in order to achieve mutual cooperation with opponents using the same strategy
but not with other opponents. Nevertheless an evolutionary process is able to
tunnel through the costs and risks associated to early defections to find more
optimal solutions, so it is not surprising in hindsight that handshaking
strategies emerge from the evolutionary training process.

A handshake requires at least one defection and there is
selective pressure to defect as few times as possible to achieve the
self-recognition mechanism. It is also unwise to defect on the first move as
some strategies additionally retaliate first round defections. So the
handshakes used by TF1, TF2, and CS are in some sense optimal. These
discoveries may have significant ramifications regarding the evolution of
cooperation and forgiveness in biological organisms such as antibacterial
resistant bacteria and social interactions between humans.

@@

This is confirmed by the performance of the zero determinant strategies. While
some do rank relatively highly for \(N=2\) (ZD-extort-4 has rank 16) this rank
does not translate to larger populations.

@@

\section{Discussion}

Training strategies to excel
at the Moran process led to the evolution of cooperation, but only with like
individuals in the case of TF1 and TF2. This may have significant implications
for human social interactions such as the evolution of ingroup/outgroup mechanisms
and other sometimes costly rituals that reinforce group behavior.

While TF1 and TF2 are competent invaders, the best invaders
in the study do not appear to employ strict handshakes, and are generally
cooperative strategies. TF3, which does not use a handshake, is a better invader
than TF1 and TF2 but not as good of a resistor. Nevertheless it was the result
of the same kind of training procceses and is a better combined invader-resistor
than the invaders that were trained previously to maximize payout.

The strategies trained to maximize payoff in head-to-head matches are generally
cooperative and are effective invaders.
Combined with the fact that handshaking strategies are stronger resisters,
this suggests that while maximizing individual payoff can lead to the evolution
of cooperation (through fixation), it is not necessarily evolutionarily stable
in the long run. A strategy with a handshaking mechanism is still capable of
invading and is more resistant to subsequent invasions. Moreover, the
best resistor of the payoff maximally trained strategies is EvolvedLookerUp1_1_1,
which always defects if the opponent defects in the first round, is effectively
employing a one-shot handshake of C. These insights also suggest that a strategy
aware of the population distribution could choose to become a handshaker at
a critical threshold and use a different strategy for invasion when in the
minority. However this information was not available to our strategies.

We did not attempt other objective functions that may serve to select for both
invasion and resistance better than training at a starting population of
$(N/2, N/2)$. Nevertheless our results suggest that there is not much room for
improvement. Any handshake more sophisticated than C necessarily involves
a defection. (A handshake consisting of a longer sequence of cooperations is
effectively a grudger.) For TF3 or EvolvedLookerUp1_1_1 to become better resistors
they need a longer, more strict handshake. But if this handshake involves
a defection then likely the invasion ability is diminished for $N > 2$: the top
invaders for larger $N$ are nice strategies that do not defect before their
opponents. This is because good invaders still need to cooperate with themselves,
and so in the absense of a handshake mechanism or knowledge of the population
distribution, a strategy must be generally cooperative. Aggressive strategies
are only effective invaders for the smallest $N$, dropping down dramatically
as the population size increases.

We did, however, attempt to evolve CS using FSM and lookup table based players,
which resulted in some very similar strategies. In particular we evolved a
lookup strategy that had a handshake of DC and played TFT with other players
after a correct handshake and defecting otherwise, which is quite close in
function to CS (full grudging is not possible with a lookup table).

Finally we note that it may be possible to achieve similar results with smaller
capacity finite state machine players.

@@

When studying evolutionary processes it is vital to consider \(N>2\) since
results for \(N=2\) cannot be used to extrapolate performance in larger
populations. This was shown both observationally in
Sections~\ref{sec:strong_invaders} and~\ref{sec:strong_resistors} but also by
considering the correlation of the ranks in different population sizes in
Section~\ref{sec:population_size}.

Memory one strategies do not perform well in general in this study. There are
no memory one strategies in the top 5 performing strategies
for \(N>3\). This is due at least partly to their lack of ability to
recognise their opponents. More sophisticated strategies
prove to be high performers for invasion: these are infinite memory strategies
which have been trained using a number of reinforcement learning algorithms.
Interestingly they have been trained to perform well in tournaments and not
Moran processes which highlights the potentially for improvement.

One of the major findings discussed in Section~\ref{sec:strong_resistors}, is
the ability of strategies with a handshake mechanism to resist invasion. This
was not only revealed for CS (a human designed strategy) but also for
two FSM strategies (TF1 and TF2) specifically trained through an evolutionary
process. In these two cases, the handshake mechanism was a product of the
evolutionary process. This has the potential to help with the understanding of
organisms with a strong resistance to invasion such as anti antibacterial
resistant bacteria \cite{Davies2010}. With the knowledge that a handshake being
likely to exist, perhaps it can be mimicked.

These findings are important for the ongoing understanding of
population dynamics and offer evidence for some of the shortcomings of low
memory which has started to be recognised by the community \cite{Hilbe2017}.

@@

% There are various areas for further work to build on this. Firstly, an analysis
% of the effect of noise would offer insights about the stability of the findings.
% It would also be possible to consider three or more types of strategy in the
% population and finally mutation would also offer an interesting dimension to
% explore.
